$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-02 paper/appendix.tex
Go over varref

>>>>

\begin{nilemma}
  \label{lm:comm-capable}
  Every input usage context $\Gamma$ of a well-typed process $\types{\gamma}{\Gamma}{P}{\Delta}$ that reduces by communicating on a channel external to it (that is, $P \reduce{\constr{external} \; i} Q$ for some $Q$) has a multiplicity of at least $\lio$ at index $i$.
\end{nilemma}

\begin{proof}
  By induction on the reduction derivation $P \reduce{\constr{external \; i}}Q$.
\end{proof}

### 2020-05-02 paper/paper.tex
Go over varref

\subsection{Structural Congruence}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{structural-congruence}

@@

\subsection{Reduction Relation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{operational-semantics}

@@

\subsection{Typing Contexts}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\label{contexts}

@@

\subsection{Typing with Leftovers}
\label{leftover-typing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We use \emph{leftover typing} \cite{Allais2018a} for our typing rules, an approach that, in addition to the usual typing context and usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} (the unused multiplicities) of the process being typed.
These leftovers can then be used as an input to another typing derivation.
This techinque presents three main advantages:
\begin{itemize}
  \item
    The transformations to the usage context (consequence of receiving and sending over a channel) are encapsulated into a variable reference judgment with both an input and an output usage context.
  \item
    The need for an extensional context splitting proof for parallel composition is eradicated.
    Such a proof would need to construct a witness $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$, where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    This could be obtained through automated search --- but it is better yet not to need it.
    Leftover typing uses the leftovers of $P$ to type $Q$.
  \item
    New theorems (like framing (\autoref{framing}), weakening (\autoref{weakening}) and strengthening (\autoref{strengthening}) can suddenly be state.
\end{itemize}

We introduce two judgments: one on variable references (\autoref{def:varref}), one on processes (\autoref{def:types}).

@@

  The \textsc{VarRef} relation presented in \autoref{fig:variable-references} is indexed by a typing context and an input and output usage context.
  In addition to these, it also keeps track of its position in the contexts, its type, and its usage annotation.

  The base case $\constr{here}$ asserts that $t$ is the most immediate type in typing context, and splits the most immediate usage annotation $x$ in the input context into the usage annotation $y$ the usage annotation $z$ in the output context.
  This is the only place in the typing system where the algebra on multiplicities is used.

  \begin{remark}
    We use $\field{\cdot-compute}$ to aliviate the user from the proof burden $\opsquared{x}{y}{z}$.
  \end{remark}

  The inductive case $\constr{there}$ appends an extra type to the typing context, and an extra usage annotation to both the input and output usage context.

@@

\begin{example}
  The following is an example that asserts that the input usage context is capable of $\li$ on a $\channel{\unit}{\li}$ type at index $\constr{suc} \; \constr{zero}$, and that consuming such $\li$ results in the stated leftovers.
  \begin{lstlisting}[mathescape,caption=Here \_ introduces an anonymous declaration that must be immediately followed by its definition.]
  _ : $\contains{\constr{[]} \comma \channel{\unit}{\li} \comma \unit} {\constr{[]} \comma \lio \comma \lio} {\constr{suc} \; \constr{zero}} {\channel{\unit}{\li}} {\li} {\constr{[]} \comma \lo \comma \lio}$
  _ = $\constr{there} \; \constr{here}$
  \end{lstlisting}
\end{example}

@@

\begin{nilemma}
  \label{lm:comm-capable}
  As an auxiliary lemma, we must establish that every input usage context $\Gamma$ of a well-typed process $\types{\gamma}{\Gamma}{P}{\Delta}$ that reduces by communicating on a channel external to it (that is, $P \reduce{\constr{external} \; i} Q$ for some $Q$) has a multiplicity of at least $\lio$ at index $i$.
\end{nilemma}

\begin{proof}
  By induction on the reduction derivation $P \reduce{\constr{external \; i}}Q$.
\end{proof}

@@

\cite{Allais2018a}
typing with leftovers
bidirectional typing rules
decidable typechecking and type inference
limited to a linear type system
framing, weakening, substitution
soundness and completeness with respect to a sequent calculus for intuitionistic linear logic

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Structural Congruence}
\label{structural-congruence}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Reduction Relation}
\label{operational-semantics}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\subsection{Typing Contexts}
\label{contexts}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\subsection{Typing with Leftovers}
\label{leftover-typing}

We use \emph{leftover typing} \cite{Allais2018a} for our typing rules, an approach that, in addition to the usual typing context and usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} (the unused multiplicities) of the process being typed.
These leftovers can then be used as an input to another typing derivation.
This techinque presents three main advantages:
\begin{itemize}
  \item
    The transformations to the usage context (consequence of receiving and sending over a channel) are encapsulated into a variable reference judgment with both an input and an output usage context.
  \item
    The need for an extensional context splitting proof for parallel composition is eradicated.
    Such a proof would need to construct a witness $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$, where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    This could be obtained through automated search --- but it is better yet not to need it.
    Leftover typing uses the leftovers of $P$ to type $Q$.
  \item
    New theorems (\textbf{framing} (\autoref{framing}), \textbf{weakening} (\autoref{weakening}) and \textbf{strengthening} (\autoref{strengthening})) can suddenly be stated.
\end{itemize}

We introduce two judgments: one on variable references (\autoref{def:varref}), one on processes (\autoref{def:types}).
Both of these are indexed by a typing context, an input usage context, and an output usage context.

@@

  The \textsc{VarRef} relation presented in \autoref{fig:variable-references} keeps track of the index, type, and usage annotation of a variable within a particular typing and usage context.
  The leftover usage context is the result of subtracting the usage annotation from the input usage context.

  The base case $\constr{here}$ splits the usage annotation $x$ into $y$ and $z$, the leftovers.
  This splitting is as per the usage algebra defined on the index of the usage annotation.

  \begin{remark}
    We use the functionality of the monoidal relation to aliviate the user from the proof burden $\opsquared{x}{y}{z}$.
  \end{remark}

@@

\begin{example}
  The following example asserts that the input $\channel{\unit}{\li}$ type at index $\constr{suc} \; \constr{zero}$ is capable of $\li$, and that consuming such $\li$ results in $\lo$.
  \begin{lstlisting}[mathescape,caption=Here the underscore \_ introduces an anonymous declaration immediately followed by its definition.]
  _ : $\contains{\constr{[]} \comma \channel{\unit}{\li} \comma \unit} {\constr{[]} \comma \lio \comma \lio} {\constr{suc} \; \constr{zero}} {\channel{\unit}{\li}} {\li} {\constr{[]} \comma \lo \comma \lio}$
  _ = $\constr{there} \; \constr{here}$
  \end{lstlisting}
\end{example}

@@

@@

This work is heavily based on the work by Allais \cite{Allais2018a}.
Allais uses leftover typing to build a bidirectional type system for the linear \lambdacalc{}.
He proves type preservation and provides a procedure that decides type checking and type inference.
He then goes on showing the soundness and completeness with respect to the sequent calculus for intuitionistic linear logic.
